---
title: 'Tidy Tuesday 2020-09-15: Government Spending on Kids'
author: 'Leo Williams'
date: |
  | Created 2020-12-04
  | Updated: `r format(Sys.time(), "%Y %m_%b %d")`
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
params:
  in_data: "submodules/tidytuesday/data/2020/2020-09-15/kids.csv"
editor_options: 
  chunk_output_type: console
---

# About

This document is a follow-along to the live Tidy Tuesday coding/analysis by David Robinson found here: [Tidy Tuesday live screencast: Analyzing government spending on kids in R](https://www.youtube.com/watch?v=dHRPrVsnNwo)

# Setup

```{r setup, message=FALSE}
# Load packages ----
library(knitr)
library(tidyverse)
library(tidytuesdayR)
library(scales)

# Set working dir for setup chunk ----
setwd(rprojroot::find_rstudio_root_file())
getwd()

# Set global knit options ----
opts_knit$set(
    # set project root dir as working dir
    root.dir = rprojroot::find_rstudio_root_file()
)
```

```{r other setup}
# copy tt data from submodule data/ to project data/ folder
theme_set(theme_classic() +
            theme(text = element_text(face = "bold"),
                  axis.text = element_text(family = "Courier")))
```

# Load Data

```{r load tt data}
tt = tt_load("2020-09-15")
kids = tt$kids
```

```{r view data}
kids
```


# Explore data

```{r explore data}
# how many distinct types of variables?
kids %>%
  distinct(variable)
```

Inspect the full [variable cookbook](https://jrosen48.github.io/tidykids/articles/tidykids-codebook.html). There are a lot of variables you could analyze.

```{r}
kids %>%
  filter(variable == "PK12ed") %>% # let's just look at public spending for K-12
  distinct(state) # 50 states + DC
```

```{r}
kids %>%
  filter(variable == "PK12ed") %>% # let's just look at public spending for K-12
  distinct(year) # over 20 years
```

```{r}
kids %>%
  filter(variable == "PK12ed") %>%
  group_by(year) %>%
  summarize(total = sum(inf_adj)) %>%
  ggplot(aes(year, total)) +
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = dollar)
```
general spending over time, but we really want to look by state.

```{r}
kids %>%
  filter(variable == "PK12ed", 
         state %in% c("New York", "Washington", "California", "Louisiana", "Texas", "Florida")) %>% # really want to look by state
  ggplot(aes(year, inf_adj_perchild, color = state)) +
  geom_line() +
  expand_limits(y = 0)
```
Some look to be kinda linear that we could look at them with an intercept term and how they increase over time

Lets pick 20 states at random
```{r}
kids %>%
  filter(variable == "PK12ed", 
         state %in% sample(unique(state), 51)) %>% # really want to look by state
  mutate(state = fct_reorder(.f = state, .x = inf_adj_perchild, .fun = max, .desc = T)) %>% # reorder plot based on max value on every states line
  ggplot(aes(year, inf_adj_perchild*1000)) + # remove color bc too many samples - replace with facet
  geom_line() +
  expand_limits(y = 0) +
  scale_y_continuous(labels = dollar) +
  facet_wrap(~ state) +
  labs(x = "Year", 
       y = "Inflation-adjusted spending per child")
```
Randomly chosen states generally show a linear trend. a plateau is noticeable around the time the financial crisis hit. lets add an indicator for that

```{r}
kids %>%
  filter(variable == "PK12ed", 
         state %in% sample(unique(state), 51)) %>% # really want to look by state
  mutate(state = fct_reorder(.f = state, .x = inf_adj_perchild, .fun = max, .desc = T)) %>% # reorder plot based on max value on every states line
  ggplot(aes(year, inf_adj_perchild*1000)) + # remove color bc too many samples - replace with facet
  geom_line() +
  geom_vline(color = "red", lty = 2, xintercept = 2009) +
  expand_limits(y = 0) +
  scale_y_continuous(labels = dollar) +
  facet_wrap(~ state) +
  labs(x = "Year", 
       y = "Inflation-adjusted spending per child")
```
lots of states' spending levels off after the crisis. we can do a PCA(?)

lets try to analyze the rate of increase rather than the absolute value of spending
```{r}
kids %>%
  filter(variable == "PK12ed") %>%
         # state %in% sample(unique(state), 51)) %>% # really want to look by state
  group_by(state) %>%
  mutate(change = inf_adj_perchild / first(inf_adj_perchild) - 1) %>%
  ungroup() %>%
  mutate(state = fct_reorder(state, change, max, .desc = T)) %>%
  arrange(state, year) %>%
  ggplot(aes(year, change)) + # remove color bc too many samples - replace with facet
  geom_line() +
  geom_vline(color = "red", lty = 2, xintercept = 2009) +
  geom_hline(yintercept = 0, color = "lightgray") +
  expand_limits(y = 0) +
  scale_y_continuous(labels = percent, breaks = c(0, 1, 2)) +
  facet_wrap(~ state) +
  labs(title = "How has K-12 spending changed per student in each state?",
       subtitle = "Red line indicates the 2009 global financial crisis",
       x = "Year", 
       y = "Increase in Inflation-adjusted spending per child relative to 1997")
```

We could apply a lot of this code to other variables in the dataset.

let's put the common plotting code into a function:
```{r}
# now separate plotting part from transformation part
plot_faceted = function(tbl, y_axis) {
    tbl %>%
    mutate(state = fct_reorder(state, {{ y_axis }}, max, .desc = T)) %>%
    # common code for plotting
    ggplot(aes(year, {{ y_axis }})) + # remove color bc too many samples - replace with facet
    geom_hline(yintercept = 0, color = "gray") +
    geom_line() +
    facet_wrap(~ state)
}

plot_change_faceted = function(tbl) {
  tbl %>%
    # need to add change variable
    group_by(state, variable) %>%
    mutate(change = inf_adj_perchild / first(inf_adj_perchild) - 1) %>%
    ungroup() %>%
    plot_faceted(change) +
    scale_y_continuous(labels = percent)
}
```

Now do the same plot but implement with the new function
```{r}
kids %>%
  filter(variable == "PK12ed") %>%
  plot_change_faceted() +
  geom_vline(color = "red", lty = 2, xintercept = 2009) +
  geom_hline(yintercept = 0, color = "lightgray") +
  labs(title = "How has K-12 spending changed per student in each state?",
       subtitle = "Red line indicates the 2009 global financial crisis",
       x = "Year", 
       y = "Increase in Inflation-adjusted spending per child relative to 1997")
```

Lets apply this to another variable:
```{r}
kids %>%
  filter(variable == "highered") %>%
  plot_change_faceted() +
  geom_vline(color = "red", lty = 2, xintercept = 2009) +
  # geom_hline(yintercept = 0, color = "lightgray") +
  labs(title = "How has higher education spending changed per student in each state?",
       subtitle = "Red line indicates the 2009 global financial crisis",
       x = "Year", 
       y = "Increase in Inflation-adjusted spending per child relative to 1997")
```

***INCOMPLETE***: just watched the rest of the screencast starting about halfway through when he starts importing the excel files.

# Ideas

We might do:
- Principal component analysis
- a choropleth (colored map by state)
- animated choropleth - showing change from 1997-2016
- what states had the fastest growing/shrinking of each of these variables
